QuickSched: Task-based parallelism with dependencies and conflicts

### 2015-10-24 paper/paper.tex
second round of corrections. still need to check the appendix and re-do the numbers.

The same decomposition was implemented using OmpSs v.\,1.99.0,
calling the kernels directly using {\tt \#pragma omp task}
annotations with the respective dependencies, and
the runtime parameters
\begin{quote}
  \tt --disable-yield --schedule=socket --cores-per-socket=16 \\--num-sockets=4
\end{quote}
\noindent Several different schedulers and parameterizations
were discussed with the authors of OmpSs and tested, with
the above settings produced the best results.

@@

Since in QuickSched the entire task structure is known explicitly
in advance, the scheduler ``knows'' that the DGEQRF tasks all
lie on the longest critical path and therefore executes them as
soon as possible.
OmpSs, does not exploit this knowledge, resulting in the less efficient
scheduling seen in \fig{QRTasks}.

@@

The Barnes-Hut tree-code \citep{ref:Barnes1986}
is an algorithm to approximate the
solution of an $N$-body problem, i.e.~computing all the
pairwise interactions between a set of $N$ particles,
in \oh{N\log N} operations, as opposed to the \oh{N^2}
naive direct computation.
The algorithm is based on a recursive octree decomposition:
Starting from a cubic cell containing all the particles,
the cell is recursively bisected along all three spatial dimensions,
resulting in eight sub-cells, until the number of particles
per cell is smaller than some limit $n_\mathsf{max}$.
The particle interactions can then be formulated recursively:
Given a particle and a set of particles in a cell,
if the particle and cell
are sufficiently well separated, the particle-cell interactions
are approximated by interacting the particle with the cell's
center of mass.
If the particle and the cell are too close, and the cell
has sub-cells, i.e.~it contained more than $n_\mathsf{max}$
particles and was split in the recursive octree decomposition,
then the particle is interacted with each of the sub-cells
recursively.
Finally, if the cell is not split, i.e.~it is a leaf cell
in the octree, then the particle is interacted with all
particles in the cell, except for the particle itself if
it is in the same cell.
This operation is performed for each particle, starting
with the root-level cell containing all the particles.

@@

These tasks can be created recursively over the cell hierarchy
as shown in the function {\tt make\_tasks} in \fig{MakeTasks}.
The function is called on the root cell with the root cell
and {\tt NULL} as its two cell parameters.
The function recurses as follows (line numbers refer to \fig{MakeTasks}:
\begin{itemize}
    \item If called with a single, split cell (lines~6--7),
        recurse over all the cell's sub-cells, and all
        pairs of the cell's sub-cells (lines~8--11),
    \item If called with a single unsplit cell (line~13),
        create a self-interaction task (line~14) as well as a particle-cell
        task on that cell (line~18),
    \item If called with two neighbouring cells and both cells
        are split (line~22),
        recurse over all pairs of sub-cells spanning
        both cells (lines~24--26), and
    \item If called with two neighbouring cells
        and one of the cells are not split, create
        a particle-particle pair task over both cells (line~29),
    \item If called with two non-neighbouring cells,
        do nothing, as these interactions
        will be computed by the particle-cell task.
\end{itemize}
\noindent where every interaction task additionally locks
the cells on which it operates (lines~17, 20, and 32--33).
In order to prevent generating
a large number of very small tasks, the task generation only recurses
if the cells contain more than a minimum number $n_\mathsf{task}$
of threads each (lines~7 and~23).

@@

\begin{figure}
    \centerline{\epsfig{file=figures/BH_scaling.pdf,width=\textwidth}}
    \caption{Strong scaling and parallel efficiency of the Barnes-Hut tree-code
        computed over 1\,000\,000 particles.
        Solving the N-Body problem takes 323\,ms, achieving 75\% parallel
        efficiency, over all 64 cores.
        For comparison, timings are shown for the same computation using
        the popular astrophysics code Gadget-2.
        The scaling for Gadget-2 (left) is shown relative to the performance of
        QuickSched, whereas the parallel efficiency (right) is computed relative
        to Gadget-2 on a single core.
        }
    \label{fig:BHResults}
\end{figure}

@@

Unlike the QR decomposition, the results scale well only to
32 cores, achieving 90\% parallel efficiency, and then
level off for increasing numbers of cores.
This, however, is not a problem of the task-based parallel
algorithm, or of QuickSched, but of the memory bandwidth
of the underlying hardware.
\fig{BHTimes} shows the accumulated cost of each task type and of 
QuickSched over the number of cores.
At 64 cores, the scheduler overheads account for only $\sim 1$\% of
the total computational cost, whereas,
as of 32 cores, the cost of both pair types grow by up to
40\%.
This is due to memory bandwidth restrictions, as
the cost of the particle-cell interaction tasks, which do significantly more
computation per memory access, only grow by up to 10\%.

\begin{figure}
    \centerline{\epsfig{file=figures/BH_times.pdf,width=0.8\textwidth}}
    \caption{Accumulated cost of each task type and of the overheads
        associated with {\tt qsched\_gettask}, summed over all cores.
        As of 32 cores, the cost of both pair interaction task
        types grow by up to 40\%.
        The cost of the particle-cell interactions, which entail significantly more
        computation per memory access, grow only by at most 10\%.
        The scheduler overheads, i.e.~{\tt qsched\_gettask},
        make up less than 1\% of the total time.}
    \label{fig:BHTimes}
\end{figure}

@@

% Acknowledgments
\section*{Acknowledgments}
The authors would like to thank Lydia Heck of the Institute for
Computational Cosmology at Durham University for providing access
to, and expertise on, the COSMA cluster used in the performance
evaluation.
This work was supported by a Durham University Seedcorn Grant.

>>>>

The same decomposition was implemented using OmpSs v.\,1.99.0,
calling the kernels directly using {\tt \#pragma omp task}
annotations with the respective dependencies, and
the runtime parameters
\begin{quote}
  \tt --disable-yield --schedule=socket --cores-per-socket=16 \\--num-sockets=4
\end{quote}
\noindent Several different schedulers and parameterizations
were discussed with the authors of OmpSs and tested, with
the above settings producing the best results.

@@

Since in QuickSched the entire task structure is known explicitly
in advance, the scheduler ``knows'' that the DGEQRF tasks all
lie on the longest critical path and therefore executes them as
soon as possible.
OmpSs does not exploit this knowledge, resulting in the less efficient
scheduling seen in \fig{QRTasks}.

@@

The Barnes-Hut tree-code \citep{ref:Barnes1986}
is an algorithm to approximate the
solution of an $N$-body problem, i.e.~computing all the
pairwise interactions between a set of $N$ particles,
in \oh{N\log N} operations, as opposed to in \oh{N^2} for the
naive direct computation.
The algorithm is based on a recursive octree decomposition:
Starting from a cubic cell containing all the particles,
the cell is recursively bisected along all three spatial dimensions,
resulting in eight sub-cells, until the number of particles
per cell is smaller than some limit $n_\mathsf{max}$.
The particle interactions can then be formulated recursively:
Given a particle and a set of particles in a cell,
if the particle and cell
are sufficiently well separated, the particle-cell interactions
are approximated by interacting the particle with the cell's
center of mass.
If the particle and the cell are too close, and the cell
has sub-cells, i.e.~it contained more than $n_\mathsf{max}$
particles and was split in the recursive octree decomposition,
then the particle is interacted with each of the sub-cells
recursively.
Finally, if the cell is not split, i.e.~it is a leaf cell
in the octree, then the particle is interacted with all
particles in the cell, except for the particle itself if
it is in the same cell.
This operation is performed for each particle, starting
with the root-level cell containing all the particles.

@@

These tasks can be created recursively over the cell hierarchy
as shown in the function {\tt make\_tasks} in \fig{MakeTasks}.
The function is called on the root cell with the root cell
and {\tt NULL} as its two cell parameters.
The function recurses as follows (line numbers refer to \fig{MakeTasks}:
\begin{itemize}
    \item If called with a single, split cell (lines~6--7),
        recurse over all the cell's sub-cells, and all
        pairs of the cell's sub-cells (lines~8--11),
    \item If called with a single unsplit cell (line~13),
        create a self-interaction task (line~14) as well as a particle-cell
        task on that cell (line~18),
    \item If called with two neighbouring cells and both cells
        are split (line~22),
        recurse over all pairs of sub-cells spanning
        both cells (lines~24--26), and
    \item If called with two neighbouring cells
        and at least one of the cells is not split, create
        a particle-particle pair task over both cells (line~29),
    \item If called with two non-neighbouring cells,
        do nothing, as these interactions
        will be computed by the particle-cell task.
\end{itemize}
\noindent Every interaction task additionally locks
the cells on which it operates (lines~17, 20, and 32--33).
In order to prevent generating
a large number of very small tasks, the task generation only recurses
if the cells contain more than a minimum number $n_\mathsf{task}$
of particles each (lines~7 and~23).

@@

\begin{figure}
    \centerline{\epsfig{file=figures/BH_scaling.pdf,width=\textwidth}}
    \caption{Strong scaling and parallel efficiency of the Barnes-Hut tree-code
        computed over 1\,000\,000 particles.
        Solving the N-Body problem takes 323\,ms, achieving 75\% parallel
        efficiency over all 64 cores.
        For comparison, timings are shown for the same computation using
        the popular astrophysics code Gadget-2.
        The scaling for Gadget-2 (left) is shown relative to the performance of
        QuickSched, whereas the parallel efficiency (right) is computed relative
        to Gadget-2 on a single core.
        }
    \label{fig:BHResults}
\end{figure}

@@

Unlike the QR decomposition, the results scale well only to
32 cores, achieving 90\% parallel efficiency, and then
level off for increasing numbers of cores.
This, however, is not a problem of the task-based parallel
algorithm, or of QuickSched, but of the memory bandwidth
of the underlying hardware.
\fig{BHTimes} shows the accumulated cost of each task type and of 
QuickSched over the number of cores.
At 64 cores, the scheduler overheads account for only $\sim 1$\% of
the total computational cost, whereas,
as of 32 cores, the cost of both pair types grow by up to
40\%.
This is due to the cache hierarchy of the AMD Opteron 6376 in which
pairs of cores share a comon 2\,MB L2 cache.
When using half the cores or less, each core has its L@ cache to
itself, whereas beyond 32 cores they are shared, resulting in more
frequent cache misses.
This cen be seen when comparing the costs of the particle-particle
interaction and particle-cell interaction tasks: while the former grow by
roughly 30\%, the latter grow by only 10\% as they do much more
computation per memory access.

\begin{figure}
    \centerline{\epsfig{file=figures/BH_times.pdf,width=0.8\textwidth}}
    \caption{Accumulated cost of each task type and of the overheads
        associated with {\tt qsched\_gettask}, summed over all cores.
        As of 32 cores, the cost of both pair interaction task
        types grow by up to 30\%.
        The cost of the particle-cell interactions, which entail significantly more
        computation per memory access, grow only by at most 10\%.
        The scheduler overheads, i.e.~{\tt qsched\_gettask},
        make up less than 1\% of the total time.}
    \label{fig:BHTimes}
\end{figure}

@@

% Acknowledgments
\section*{Acknowledgments}
The authors would like to thank Tom Theuns and Richard Bowers of the
Institute for Computational Cosmology at Durham University for the
helpful discussions.
This work was supported by a Durham University Seedcorn Grant
number 21.12.080130 from
which the hardware used in the experiments was purchased.
